---
title: "Lecture21Oct2025"
author: "Maya Powell"
date: "October 28 2025"
date-modified: today
format:
  html:
    toc: true
#    theme: darkly
editor: visual
warning: false
---

## Working with Factors

#### Load libraries and set up data

```{r}
library(tidyverse)
library(here)

set.seed(8)

df <- tibble(
  a = rnorm(10), # draws 10 random values from a normal distribution
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
head(df)
```

Mutate data - rescale every column individually - booo

```{r}
df<-df %>%
  mutate(a = (a-min(a, na.rm = TRUE))/(max(a, na.rm = TRUE)-min(a, na.rm = TRUE)))

df<-df %>%
  mutate(a = (a-min(a, na.rm = TRUE))/(max(a, na.rm = TRUE)-min(a, na.rm = TRUE)),
         b = (b-min(b, na.rm = TRUE))/(max(a, na.rm = TRUE)-min(b, na.rm = TRUE)), #UH OH MISTAKE!! don't need to copy/paste - instead lets write a function
         c = (c-min(c, na.rm = TRUE))/(max(c, na.rm = TRUE)-min(c, na.rm = TRUE)),
         d = (d-min(d, na.rm = TRUE))/(max(d, na.rm = TRUE)-min(d, na.rm = TRUE)))
```

#### Write functions

Instead let's write a function to help us do this! yayyyy

You can do mutate_all to do all lines, mutate_at for certain lines, or mutate_if for conditional ones

```{r}
rescale01 <- function(x) {
  value<-(x-min(x, na.rm = TRUE))/(max(x, na.rm = TRUE)-min(x, na.rm = TRUE))
  return(value)
}

df %>%
  mutate(a = rescale01(a),
         b = rescale01(b),
         c = rescale01(c),
         d = rescale01(d))

df %>%
  mutate_all(rescale01)
```

Make a function to convert °C to °F

```{r}
F_to_C <- function(temp_F) { 
  temp_C <- (temp_F - 32) * 5 / 9 
  return(temp_C)
}

F_to_C(32)
F_to_C(212)
```

Make sure to always test your function for an expected output

Now make one for °C to K

```{r}
C_to_K <- function(temp_C) { 
  temp_K <- (temp_C + 273.15) 
  return(temp_K)
}

C_to_K(0)
C_to_K(100)
```

#### Make plots into a function

First make a line plot

```{r}
library(palmerpenguins)
library(PNWColors) # for the PNW color palette 
# you may need to install the PNWColors library if you haven't used it yet
pal<-pnw_palette("Lake",3, type = "discrete") # my color palette 
ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm, color = island))+
  geom_point()+
  geom_smooth(method = "lm")+ # add a linear model
  scale_color_manual("Island", values=pal)+   # use pretty colors and another example of how to manually change the legend title for colors
  theme_bw()
```

Now make it into a function

here we have to use a curly-curly:\
{rlang} uses what is literally called a"curly-curly" {{}} to help us assign variables that are column names in dataframes.

Here we add curly-curlies to the column names

```{r}
myplot<-function(data, x, y){ 
pal<-pnw_palette("Lake",3, type = "discrete") # my color palette 
ggplot(data, aes(x = {{x}}, y = {{y}} , color = island))+
  geom_point()+
  geom_smooth(method = "lm")+ # add a linear model
  scale_color_manual("Island", values=pal)+   # use pretty colors and change the legend title
  theme_bw()
}

#make a plot
myplot(data = penguins, x = body_mass_g, y = bill_length_mm)

#now make another plot
#just make sure it's continuous data
myplot(data = penguins, x = body_mass_g, y = flipper_length_mm)

```

Adding defaults

Make it specific to the dataset

```{r}
myplot<-function(data = penguins, x, y){
pal<-pnw_palette("Lake",3, type = "discrete") # my color palette 
ggplot(data, aes(x = {{x}}, y = {{y}} , color = island))+
  geom_point()+
  geom_smooth(method = "lm")+ # add a linear model
  scale_color_manual("Island", values=pal)+   # use pretty colors and change the legend title
  theme_bw()
}

myplot(x = body_mass_g, y = flipper_length_mm) #this assumes that data is penguins dataset, you can still change the dataset but this makes it easier for you

#you can still layer on to this like any other ggplot code to add labels here etc
myplot(x = body_mass_g, y = flipper_length_mm)+
  labs(x = "Body mass (g)",
       y = "Flipper length (mm)")
```

#### If-else statements in functions

```{r}
a <- 4
b <- 5

if (a > b) { # my question
  f <- 20 # if it is true give me answer 1
    } else { # else give me answer 2
  f <- 10
    }

f

```

Plotting with if-else statements

Using if\...else\... we can make our function even more flexible. Let's say we want the option of adding the geom_smooth lines. We can create a variable that if set to TRUE add the geom_smooth, otherwise print without.

First add a new argument for lines and make the default TRUE for ease

```{r}
myplot<-function(data = penguins, x, y ,lines=TRUE ){ # add new argument for lines
pal<-pnw_palette("Lake",3, type = "discrete") # my color palette 
ggplot(data, aes(x = {{x}}, y = {{y}} , color = island))+
  geom_point()+
  geom_smooth(method = "lm")+ # add a linear model
  scale_color_manual("Island", values=pal)+   # use pretty colors and change the legend title
  theme_bw()
}

#Next, add an if-else statement

myplot<-function(data = penguins, x, y, lines=TRUE ){ # add new argument for lines
pal<-pnw_palette("Lake",3, type = "discrete") # my color palette 
if(lines==TRUE){
ggplot(data, aes(x = {{x}}, y = {{y}} , color = island))+
  geom_point()+
  geom_smooth(method = "lm")+ # add a linear model
  scale_color_manual("Island", values=pal)+   # use pretty colors and change the legend title
  theme_bw()
}
else{
ggplot(data, aes(x = {{x}}, y = {{y}} , color = island))+
  geom_point()+
  scale_color_manual("Island", values=pal)+   # use pretty colors and change the legend title
  theme_bw()
}
}

#and test it out
myplot(x = body_mass_g, y = flipper_length_mm)

myplot(x = body_mass_g, y = flipper_length_mm, lines = FALSE)

```

A new fun r package

```{r}
devtools::install_github("itsrainingdata/emokid")
library(emokid)
iamsad()
iamlesssad()
mymood()
```
